---
title: "scRNA Benchmark Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    self_contained: false
params:
  metrics: ~
  timings: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = c("png", "pdf"), dpi = 300
)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(readr)
  library(scales)
})
```

```{r load-data}
metrics <- params$metrics
timings <- params$timings

keys_m <- names(metrics)[names(metrics) %in% names(timings)]
keys_t <- names(timings)[names(timings) %in% names(metrics)]

stopifnot(identical(keys_m, keys_t))

stopifnot(identical(
  metrics %>% select(all_of(keys_m)),
  timings %>% select(all_of(keys_t)),
))

timing_value_cols <- setdiff(names(timings), keys_m)
timings <- timings %>%
  mutate(across(timing_value_cols, ~ readr::parse_number(as.character(.x)))) %>%
  rename_with(~ paste0("timings_", .x), all_of(timing_value_cols))

metrics_all <- bind_cols(metrics, select(timings, -all_of(keys_m)))

metrics_all <- metrics_all %>%
  mutate(
    cluster = as.integer(cluster),
    n_comp = as.integer(n_comp),
    n_neig = as.integer(n_neig),
    n_hvg = as.integer(n_hvg),
    dataset = as.factor(dataset),
    method = as.factor(method),
    filtering = as.factor(filtering),
    module = as.factor(module)
  )


metrics_long <- metrics_all %>%
  pivot_longer(
    cols = c(
      agree_ari_leiden, agree_ari_louvain,
      struc_silhouette_leiden, struc_silhouette_louvain,
      resolution_leiden, resolution_louvain,
      n_clusters_leiden, n_clusters_louvain
    ),
    names_to = c(".value", "algorithm"),
    names_pattern =
      "^(agree_ari|struc_silhouette|n_clusters|resolution)_(leiden|louvain)$"
  ) %>%
  rename(
    ari = agree_ari,
    silhouette = struc_silhouette
  ) %>%
  mutate(
    n_clusters = as.integer(n_clusters),
    resolution = as.numeric(resolution),
    algorithm = as.factor(algorithm)
  )

metrics_new <- metrics_all %>%
  filter(module %in% c("pipelines-r", "pipelines-py"))
metrics_long_new <- metrics_long %>%
  filter(module %in% c("pipelines-r", "pipelines-py"))

metrics_all %>% glimpse()
metrics_long %>% glimpse()
```

# ARI vs number of clusters

These curves show ARI as a function of the number of clusters, grouped by method and filtering mode, and faceted by dataset and algorithm.

We only include runs with manual filtering, `n_comp = 50`, `n_neig = 20`, and `n_hvg = 1000`.

```{r ari-curves, fig.height=6, fig.width=10}
ari_curves <- metrics_long_new %>%
  filter(
    filtering == "manual",
    n_comp == 50,
    n_neig == 20,
    n_hvg == 1000
  ) %>%
  group_by(dataset, method, filtering, algorithm) %>%
  arrange(n_clusters, .by_group = TRUE)

ggplot(
  ari_curves,
  aes(x = n_clusters, y = ari, color = method)
) +
  geom_point(size = 2, alpha = 0.8) +
  geom_path(alpha = 0.8) +
  facet_grid(algorithm ~ dataset, scales = "free_x") +
  labs(
    x = "Number of clusters",
    y = "ARI",
    color = "Method"
  ) +
  theme_minimal(base_size = 12)
```

# Resolution to cluster count mapping

This view shows how resolution translates to cluster count for each algorithm.

```{r resolution-vs-clusters, fig.height=6, fig.width=10}
resolution_map <- metrics_long_new %>%
  group_by(dataset, method, filtering, algorithm, n_clusters) %>%
  summarize(resolution = mean(resolution, na.rm = TRUE), .groups = "drop")

ggplot(
  resolution_map,
  aes(x = resolution, y = n_clusters, color = method, linetype = filtering)
) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  facet_grid(algorithm ~ dataset, scales = "free_y") +
  labs(
    x = "Resolution",
    y = "Number of clusters",
    color = "Method",
    linetype = "Filter mode"
  ) +
  theme_minimal(base_size = 12)
```

# Robustness: ARI distribution across parameter settings

This shows how stable each method is across parameter choices.

```{r robustness-ari, fig.height=6, fig.width=12}
ari_robust <- metrics_long_new %>%
  filter(filtering == "manual")

ggplot(
  ari_robust,
  aes(x = method, y = ari, fill = method)
) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.size = 0.4, alpha = 0.9) +
  facet_grid(algorithm ~ dataset, scales = "free_y") +
  labs(
    x = "Method",
    y = "ARI"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )
```

# Tradeoffs: agreement vs structure

Scatter of ARI (agreement) vs silhouette (structure) to see tradeoffs per method.

```{r tradeoff-ari-silhouette, fig.height=6, fig.width=10}
tradeoff_df <- metrics_long_new %>% filter(filtering == "manual")

ggplot(
  tradeoff_df,
  aes(x = silhouette, y = ari, color = method)
) +
  geom_point(alpha = 0.6, size = 1.4) +
  facet_grid(algorithm ~ dataset, scales = "free") +
  labs(
    x = "Silhouette",
    y = "ARI",
    color = "Method"
  ) +
  theme_minimal(base_size = 12)
```

## rapids-singlecell version comparison (0.13.5 vs 0.14.1)

```{r rapids-version-comparison, fig.height=6, fig.width=10}
rapids_versions <- metrics_long %>%
  filter(
    method == "rapids",
    module %in% c("pipelines-py", "pipelines-py-old"),
    filtering == "manual",
    n_comp == 30,
    n_neig == 20,
    n_hvg == 1000
  ) %>%
  mutate(
    module = recode(
      module,
      "pipelines-py" = "v0.14.1",
      "pipelines-py-old" = "v0.13.5"
    )
  )

ggplot(
  rapids_versions,
  aes(x = n_clusters, y = ari, color = module)
) +
  geom_point(size = 2, alpha = 0.85) +
  geom_line(alpha = 0.85) +
  facet_grid(algorithm ~ dataset, scales = "free_x") +
  labs(
    x = "Number of clusters",
    y = "ARI",
    color = "rapids-singlecell version"
  ) +
  theme_minimal(base_size = 12)
```

# Timings overview

We restrict to runs done with manual filtering.

```{r timings-summary, fig.height=6, fig.width=10}
timing_cols <- grep("^timings_", names(metrics_new), value = TRUE)

timings_long <- metrics_new %>%
  pivot_longer(
    cols = all_of(timing_cols),
    names_to = "stage",
    values_to = "seconds"
  ) %>%
  mutate(stage = sub("^timings_", "", stage))

timings_summary <- timings_long %>%
  filter(filtering == "manual") %>%
  group_by(dataset, method, stage) %>%
  summarize(mean_seconds = mean(seconds, na.rm = TRUE), .groups = "drop")

ggplot(timings_summary, aes(x = dataset, y = mean_seconds, fill = method)) +
  geom_col(position = "dodge") +
  facet_wrap(~stage, scales = "free_y") +
  scale_y_continuous(labels = label_number()) +
  labs(
    x = "Dataset",
    y = "Mean seconds",
    fill = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
